---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# intraclass

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/jmgirard/intraclass/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/jmgirard/intraclass/actions/workflows/check-standard.yaml)
[![Codecov test coverage](https://codecov.io/gh/jmgirard/intraclass/graph/badge.svg)](https://app.codecov.io/gh/jmgirard/intraclass)
<!-- badges: end -->

**intraclass** computes interrater-reliability **intraclass correlation
coefficients (ICCs)** within the generalizability-theory framework, using
**modern variance-component estimation** (linear mixed models) rather than the
classical ANOVA / mean-squares approach.

It aims to (1) fit variance components with modern engines, (2) compute the
*correct* ICC for a stated design with proper (boundary-aware Monte-Carlo)
interval estimation, (3) handle imbalanced, incomplete, and multilevel designs,
and (4) help you decide **which ICC to choose, and why** — the docs and website
are a place to learn ICC best practice, not just call functions.

> [!NOTE]
> This package is approaching its first release. The full interrater-reliability
> ICC family is implemented — two-way (absolute agreement vs. consistency, single
> vs. average, random vs. fixed raters) and one-way designs, imbalanced and
> incomplete (missing-cell) data, and multilevel designs (subject vs. cluster
> level, with raters crossed with or nested in clusters/subjects) — each with
> boundary-aware Monte-Carlo intervals. Fits run on mixed-model engines
> (`glmmTMB`, `lme4`) or an SEM engine (`lavaan`). A
> [*Choosing an ICC*](https://jmgirard.github.io/intraclass/articles/choosing-an-icc.html)
> decision guide, the `choose_icc()` helper, D-study projection to other rater
> counts (`d_study()`), and `autoplot()` plots have all shipped. A Bayesian engine
> is on the roadmap; see `project/MILESTONES.md`.

## Installation

You can install the development version from [GitHub](https://github.com/jmgirard/intraclass) with:

``` r
# install.packages("pak")
pak::pak("jmgirard/intraclass")
```

The base install is light — only `glmmTMB`, `cli`, `rlang`, and `generics`.
Optional Bayesian and SEM engines live in `Suggests`.

## Example

`ratings` is the classic Shrout & Fleiss (1979) example, shipped with the
package. The defaults give the two-way random, absolute-agreement `ICC(A,1)` and
`ICC(A,k)` with a reproducible Monte-Carlo interval:

```{r example}
library(intraclass)

icc(ratings, score, subject, rater, seed = 2024)
```

Which coefficient you want — agreement vs. consistency, single vs. average, fixed
vs. random raters, complete vs. incomplete — is a real modelling decision, and each
is an argument to `icc()`. Not sure which to report? `choose_icc()` walks the
[*Choosing an ICC*](https://jmgirard.github.io/intraclass/articles/choosing-an-icc.html)
decision tree and hands back the coefficient, the reasoning, and the exact call to
run — no data or fitting required:

```{r choose}
choose_icc(model = "twoway", type = "consistency", unit = "average", raters = "random")
```

For **multilevel** data — subjects nested in clusters (pupils in classrooms,
patients in clinics) — pass a `cluster` column and `icc()` reports subject-level and
cluster-level reliability separately:

```{r multilevel}
set.seed(1)
grid <- expand.grid(pupil = 1:5, classroom = 1:12, rater = 1:4)
grid$score <- with(grid,
  10 + rnorm(12, sd = 1.2)[classroom] +
    rnorm(60, sd = 0.6)[(classroom - 1) * 5 + pupil] +
    rnorm(4, sd = 0.4)[rater] + rnorm(nrow(grid), sd = 0.7))
school <- data.frame(
  pupil = factor(paste(grid$classroom, grid$pupil, sep = "_")),
  classroom = factor(grid$classroom),
  rater = factor(grid$rater),
  score = grid$score)

icc(school, score, subject = pupil, rater = rater, cluster = classroom, seed = 2024)
```

The [*Advanced*](https://jmgirard.github.io/intraclass/articles/advanced.html)
article works through incomplete and nested multilevel designs, the estimation
engines, the `autoplot()` plots, and `choose_icc()` in full.

## Related work

Classical tools (`psych`, `irr`, `irrNA`, `irrICC`, `ICCDesign`) are ANOVA /
mean-squares based and mostly assume balanced data; model-based tools
(`performance::icc`, `gtheory`, `misty`) extract a variance-partition coefficient
but not the full interrater-reliability ICC family, the error-variance framing, or
a selection framework. **intraclass** fills that gap with mixed-model estimation,
Monte-Carlo confidence intervals, and decision guidance (following ten Hove,
Jorgensen & van der Ark).
