---
title: "Advanced: imbalanced & multilevel designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced: imbalanced & multilevel designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Beyond the balanced case

This article covers where the mixed-model approach earns its keep:

- **Multilevel ICCs** — subject-level vs. cluster-level coefficients (below).
- **Engine choice** — linear mixed models vs. SEM vs. Bayesian, and when each
  matters (a forthcoming milestone).
- **Confidence-interval methods** — Monte-Carlo (the default) and its
  alternatives, especially near the zero-rater-variance boundary.

Incomplete and imbalanced designs are already supported; the
[*Choosing an ICC*](choosing-an-icc.html) article works through a connected
incomplete design, the effective number of raters `k_eff`, and why fixed and
random raters diverge once cells are missing.

## How many raters do I need? A D-study

`ICC(*,1)` is the reliability of a *single* rater and `ICC(*,k)` the reliability
of the mean of the `k` raters you actually used. A **decision (D-) study** asks a
forward-looking question: *how reliable would the mean of some other number of
raters `m` be?* In generalizability theory the absolute-agreement ICC is the
dependability coefficient, and projecting it to `m` raters is just a change of the
averaging divisor,

$$\Phi(m) = \frac{\sigma^2_s}{\sigma^2_s + (\sigma^2_r + \sigma^2_{res}) / m},$$

so `d_study()` reuses the fit you already have — no refitting.

```{r dstudy, eval = requireNamespace("glmmTMB", quietly = TRUE)}
library(intraclass)

fit <- icc(ratings, score, subject, rater, seed = 1)
proj <- d_study(fit, m = 1:8)
proj
```

Reliability climbs with more raters but with diminishing returns, and the
projection is anchored to what you observed: at `m = 4` (the number of raters in
`ratings`) `Φ(m)` is exactly the `ICC(A,k)` you would get from `icc()` directly.
For a one-off value you can also ask `icc()` for it inline with a numeric `unit`,
e.g. `unit = c("single", "average", 6)`.

Read as a curve, this is the classic "how many raters?" picture — plot it with
`autoplot()` (which needs **ggplot2**):

```{r dstudy-plot, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("glmmTMB", quietly = TRUE), fig.alt = "Projected reliability rising with the number of raters, with a Monte-Carlo interval band."}
library(ggplot2)
autoplot(d_study(fit, m = 1:12))
```

**Projection is extrapolation.** The rater variance $\sigma^2_r$ is estimated from
only as many raters as you observed, so projecting far beyond that design leans
hard on that estimate. The Monte-Carlo interval widens honestly to reflect this
rather than pretending to a precision it lacks — and projecting absolute agreement
is refused for *fixed* raters, where there is no wider rater universe to
generalize to (use `raters = "random"`).

## Multilevel ICCs: subject vs. cluster level

Everything above treats the **subject** as the object of measurement. But subjects
are often nested in higher-level **clusters** — pupils in classrooms, patients in
clinics — and then "reliability" splits in two:

- **Subject level** (within-cluster): how reliably do raters distinguish
  *subjects within a cluster*?
- **Cluster level** (between-cluster): how reliably do raters distinguish
  *cluster means*?

Ignoring the nesting *conflates* these and biases both (ten Hove, Jorgensen & van
der Ark, 2022). Passing a `cluster` column to `icc()` fits the multilevel model
and reports each level separately.

Consider pupils nested in classrooms, each pupil rated by the same panel of
raters. We simulate a design with substantial classroom-level signal but modest
within-classroom differences:

```{r ml-data}
library(intraclass)

set.seed(2025)
n_class <- 16
n_pupil <- 5
n_rater <- 4
grid <- expand.grid(
  pupil = seq_len(n_pupil),
  classroom = seq_len(n_class),
  rater = seq_len(n_rater)
)
class_effect <- rnorm(n_class, sd = 1.3)[grid$classroom]
pupil_effect <- rnorm(n_class * n_pupil, sd = 0.6)[
  (grid$classroom - 1) * n_pupil + grid$pupil
]
rater_effect <- rnorm(n_rater, sd = 0.4)[grid$rater]
school <- data.frame(
  classroom = factor(grid$classroom),
  pupil = factor(paste(grid$classroom, grid$pupil, sep = "_")),
  rater = factor(grid$rater),
  score = 10 + class_effect + pupil_effect + rater_effect +
    rnorm(nrow(grid), sd = 0.7)
)
```

```{r ml-fit, eval = requireNamespace("glmmTMB", quietly = TRUE)}
icc(school, score, subject = pupil, rater = rater, cluster = classroom, seed = 1)
```

Both levels come back in one call. Here the **cluster-level** ICC is the higher
of the two: raters agree more about which *classrooms* score high than about which
*pupils within a classroom* do — exactly the pattern you would expect when most of
the true variation lives between classrooms. Which number you report depends on the
decision you will make: a classroom-level intervention cares about the cluster-level
reliability, a pupil-level one about the subject level. Request just one with
`level = "subject"` or `level = "cluster"`.

### When raters are nested

The classroom example above has every rater rate every pupil in every classroom,
so raters are **crossed** with clusters (ten Hove et al.'s Design 1). Two other
layouts are common, and `icc()` **infers which one you have** from the data — you
never declare it:

- **Raters nested in clusters** (Design 2): each classroom has its *own* panel of
  raters. There is then no between-cluster reliability to report — a cluster-level
  ICC needs the *same* raters spanning clusters — so `icc()` returns the subject
  level only.
- **Raters nested in subjects** (Design 3): each pupil is rated by their *own*
  raters. Now systematic rater differences cannot be separated from residual error
  at all, so this is a multilevel *one-way* design: it reports agreement-only
  `ICC(1)` / `ICC(k)`, the clustered analogue of `model = "oneway"`.

Take the same classrooms but give each one its own raters (Design 2):

```{r ml-nested-clusters, eval = requireNamespace("glmmTMB", quietly = TRUE)}
school_d2 <- school
school_d2$rater <- factor(paste(school_d2$classroom, school_d2$rater, sep = "_"))
icc(school_d2, score, subject = pupil, rater = rater, cluster = classroom, seed = 1)
```

The header now reads *raters nested in clusters* and only the subject level comes
back. If instead each *pupil* has their own raters, the design is a multilevel
one-way (Design 3):

```{r ml-nested-subjects, eval = requireNamespace("glmmTMB", quietly = TRUE)}
school_d3 <- school
school_d3$rater <- factor(paste(school_d3$pupil, school_d3$rater, sep = "_"))
icc(school_d3, score, subject = pupil, rater = rater, cluster = classroom, seed = 1)
```

Here the coefficients are labelled `ICC(1)` / `ICC(k)` and `type` no longer
applies — with each pupil's raters unique, there is no rater main effect to keep
in or drop from the error term. A layout that is neither cleanly crossed nor
cleanly nested (some raters shared across clusters, some not) raises an
informative error rather than guessing at a model.

### Incomplete (ragged) multilevel designs

Just as in the single-level case ([*Beyond the balanced case*](#beyond-the-balanced-case)),
the **crossed** design (Design 1) does not need every pupil rated by every rater.
The mixed model estimates the variance components from whatever cells are present,
so a ragged classroom design is handled directly. Drop a fifth of the ratings from
the `school` data at random:

```{r ml-incomplete-data}
set.seed(11)
school_ragged <- school[-sample(nrow(school), round(0.2 * nrow(school))), ]
```

At the **subject** level both agreement and consistency come back, and — exactly as
in the single-level incomplete case — `ICC(*,k)` averages over the *effective* number
of ratings per pupil (`k_eff`, the harmonic mean), which is below the full panel size
of `r n_rater`:

```{r ml-incomplete-subject, eval = requireNamespace("glmmTMB", quietly = TRUE)}
icc(school_ragged, score, subject = pupil, rater = rater, cluster = classroom,
  level = "subject", seed = 1)
```

The header now reads *incomplete*, and the report names the effective `k` so the
divisor is never a black box.

At the **cluster** level, the single-rater `ICC(c,1)` is available on ragged data
(it needs no averaging divisor). Request it with `unit = "single"`:

```{r ml-incomplete-cluster, eval = requireNamespace("glmmTMB", quietly = TRUE)}
icc(school_ragged, score, subject = pupil, rater = rater, cluster = classroom,
  level = "cluster", type = "consistency", unit = "single", seed = 1)
```

Two things are deliberately fenced off on incomplete data, each with a clear error
rather than a silently wrong number. The **averaged** cluster-level `ICC(c,k)` is not
yet supported: the effective number of raters behind a ragged *cluster* mean is a
per-cluster quantity still being validated, distinct from the per-pupil `k_eff`. And
when missing cells make the crossing pattern **ambiguous** — some raters happen to
appear in only one classroom, so the design could be read as crossed *or* nested —
`icc()` does not guess; you resolve it by declaring `design = "crossed"` (validated
against the data), or the abort points you at the nested reading.

Multilevel support currently covers random raters on: **crossed** designs (Design 1),
complete or **incomplete**, at the subject level and — for `ICC(c,1)` — the cluster
level; and **nested** designs (Designs 2 and 3, subject level), which still require
complete data. The averaged cluster-level `ICC(c,k)` on incomplete data, incomplete
nested designs, and fixed raters are planned for later milestones.

## Choosing an estimation engine

By default `icc()` fits the variance components with **glmmTMB**. You can instead
request **lme4** with `engine = "lme4"` for the random two-way design. Both are
REML mixed-model fits of the same model, so on a given dataset they return the
same coefficients to numerical tolerance — the choice is about the fitting
backend, not the estimand.

```{r engine, eval = requireNamespace("lme4", quietly = TRUE) && requireNamespace("merDeriv", quietly = TRUE)}
glmmtmb <- tidy(icc(ratings, score, subject, rater, engine = "glmmTMB", seed = 1))
lme4 <- tidy(icc(ratings, score, subject, rater, engine = "lme4", seed = 1))
data.frame(
  index = glmmtmb$index,
  glmmTMB = round(glmmtmb$estimate, 4),
  lme4 = round(lme4$estimate, 4)
)
```

The two point estimates agree to well within rounding, and their Monte-Carlo
intervals coincide to about `0.01`: the lme4 interval is built from the parameter
covariance supplied by the **merDeriv** package, transformed onto the same
boundary-aware log-scale glmmTMB uses. glmmTMB remains the recommended default —
it is the one required dependency and it is robust when a variance component sits
exactly at the zero boundary, where the lme4 route cannot form an interval and
directs you back to glmmTMB. lme4 for fixed-rater and multilevel designs is
planned for a later milestone.

### A structural-equation engine (`lavaan`)

`engine = "lavaan"` fits the same design as a **structural equation model** — a
common-factor generalizability model in the sense of Jorgensen (2021) — for the
complete, balanced random two-way design. Unlike lme4, this is not just a different
backend for the *same* estimator: it matters which coefficient you ask for.

```{r lavaan, eval = requireNamespace("lavaan", quietly = TRUE)}
axes <- expand.grid(
  type = c("agreement", "consistency"),
  unit = c("single", "average"),
  stringsAsFactors = FALSE
)
compare <- do.call(rbind, Map(function(type, unit) {
  g <- icc(ratings, score, subject, rater, type = type, unit = unit,
           engine = "glmmTMB", seed = 1)
  l <- icc(ratings, score, subject, rater, type = type, unit = unit,
           engine = "lavaan", seed = 1)
  data.frame(
    index = tidy(g)$index,
    glmmTMB = round(tidy(g)$estimate, 4),
    lavaan = round(tidy(l)$estimate, 4)
  )
}, axes$type, axes$unit))
compare[!duplicated(compare$index), ]
```

**Consistency** coefficients are a ratio of the subject and residual variances, so
the SEM returns them identically to the mixed model. **Absolute agreement** is
different. The SEM has no random rater effect to estimate — a rater is a single
column, so its effect lives in the column *means*. Following Jorgensen (2021), the
rater variance is recovered from the mean structure as the variance of the estimated
indicator intercepts. This **indicator-mean estimator** is a genuinely different
estimator of the rater variance than the mixed model's random effect: the two are
asymptotically equivalent and match conventional generalizability-theory software
(GENOVA, `gtheory`) closely on real data [Vispoel et al. 2022], but on a small
design they differ by a modest amount — here `ICC(A,1)` is about `0.284` from lavaan
versus `0.290` from the mixed model, because the raw variance of only four estimated
rater means carries small-sample noise the mixed model shrinks away.

Which is "right"? Neither is wrong — they are two defensible estimators of the same
population quantity. Use `"glmmTMB"` (the default) if you want the mixed-model
random-rater estimate and its wider, generalize-to-new-raters interval; reach for
`"lavaan"` if you are working inside an SEM generalizability-theory workflow and want
results comparable to that literature. The SEM engine currently covers the complete,
balanced random two-way design; one-way, fixed-rater, multilevel, and incomplete
designs are directed to the mixed-model engines.
