---
title: "Advanced: imbalanced & multilevel designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced: imbalanced & multilevel designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Beyond the balanced case

This article will cover where the mixed-model approach earns its keep:

- **Multilevel ICCs** — subject-level vs. cluster-level coefficients.
- **Engine choice** — linear mixed models vs. SEM vs. Bayesian, and when each
  matters.
- **Confidence-interval methods** — Monte-Carlo (the default) and its
  alternatives, especially near the zero-rater-variance boundary.

Incomplete and imbalanced designs are already supported; the
[*Choosing an ICC*](choosing-an-icc.html) article works through a connected
incomplete design, the effective number of raters `k_eff`, and why fixed and
random raters diverge once cells are missing.

## How many raters do I need? A D-study

`ICC(*,1)` is the reliability of a *single* rater and `ICC(*,k)` the reliability
of the mean of the `k` raters you actually used. A **decision (D-) study** asks a
forward-looking question: *how reliable would the mean of some other number of
raters `m` be?* In generalizability theory the absolute-agreement ICC is the
dependability coefficient, and projecting it to `m` raters is just a change of the
averaging divisor,

$$\Phi(m) = \frac{\sigma^2_s}{\sigma^2_s + (\sigma^2_r + \sigma^2_{res}) / m},$$

so `d_study()` reuses the fit you already have — no refitting.

```{r dstudy, eval = requireNamespace("glmmTMB", quietly = TRUE)}
library(intraclass)

fit <- icc(ratings, score, subject, rater, seed = 1)
proj <- d_study(fit, m = 1:8)
proj
```

Reliability climbs with more raters but with diminishing returns, and the
projection is anchored to what you observed: at `m = 4` (the number of raters in
`ratings`) `Φ(m)` is exactly the `ICC(A,k)` you would get from `icc()` directly.
For a one-off value you can also ask `icc()` for it inline with a numeric `unit`,
e.g. `unit = c("single", "average", 6)`.

Read as a curve, this is the classic "how many raters?" picture — plot it with
`autoplot()` (which needs **ggplot2**):

```{r dstudy-plot, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("glmmTMB", quietly = TRUE), fig.alt = "Projected reliability rising with the number of raters, with a Monte-Carlo interval band."}
library(ggplot2)
autoplot(d_study(fit, m = 1:12))
```

**Projection is extrapolation.** The rater variance $\sigma^2_r$ is estimated from
only as many raters as you observed, so projecting far beyond that design leans
hard on that estimate. The Monte-Carlo interval widens honestly to reflect this
rather than pretending to a precision it lacks — and projecting absolute agreement
is refused for *fixed* raters, where there is no wider rater universe to
generalize to (use `raters = "random"`).

> The multilevel and alternative-engine material below is a placeholder, filled in
> from **Milestone 5** onward (`project/MILESTONES.md`).
