---
title: "Advanced: imbalanced & multilevel designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced: imbalanced & multilevel designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Beyond the balanced case

This article covers where the mixed-model approach earns its keep:

- **Multilevel ICCs** — subject-level vs. cluster-level coefficients (below).
- **Engine choice** — linear mixed models vs. SEM vs. Bayesian, and when each
  matters (a forthcoming milestone).
- **Confidence-interval methods** — Monte-Carlo (the default) and its
  alternatives, especially near the zero-rater-variance boundary.

Incomplete and imbalanced designs are already supported; the
[*Choosing an ICC*](choosing-an-icc.html) article works through a connected
incomplete design, the effective number of raters `k_eff`, and why fixed and
random raters diverge once cells are missing.

## How many raters do I need? A D-study

`ICC(*,1)` is the reliability of a *single* rater and `ICC(*,k)` the reliability
of the mean of the `k` raters you actually used. A **decision (D-) study** asks a
forward-looking question: *how reliable would the mean of some other number of
raters `m` be?* In generalizability theory the absolute-agreement ICC is the
dependability coefficient, and projecting it to `m` raters is just a change of the
averaging divisor,

$$\Phi(m) = \frac{\sigma^2_s}{\sigma^2_s + (\sigma^2_r + \sigma^2_{res}) / m},$$

so `d_study()` reuses the fit you already have — no refitting.

```{r dstudy, eval = requireNamespace("glmmTMB", quietly = TRUE)}
library(intraclass)

fit <- icc(ratings, score, subject, rater, seed = 1)
proj <- d_study(fit, m = 1:8)
proj
```

Reliability climbs with more raters but with diminishing returns, and the
projection is anchored to what you observed: at `m = 4` (the number of raters in
`ratings`) `Φ(m)` is exactly the `ICC(A,k)` you would get from `icc()` directly.
For a one-off value you can also ask `icc()` for it inline with a numeric `unit`,
e.g. `unit = c("single", "average", 6)`.

Read as a curve, this is the classic "how many raters?" picture — plot it with
`autoplot()` (which needs **ggplot2**):

```{r dstudy-plot, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("glmmTMB", quietly = TRUE), fig.alt = "Projected reliability rising with the number of raters, with a Monte-Carlo interval band."}
library(ggplot2)
autoplot(d_study(fit, m = 1:12))
```

**Projection is extrapolation.** The rater variance $\sigma^2_r$ is estimated from
only as many raters as you observed, so projecting far beyond that design leans
hard on that estimate. The Monte-Carlo interval widens honestly to reflect this
rather than pretending to a precision it lacks — and projecting absolute agreement
is refused for *fixed* raters, where there is no wider rater universe to
generalize to (use `raters = "random"`).

## Multilevel ICCs: subject vs. cluster level

Everything above treats the **subject** as the object of measurement. But subjects
are often nested in higher-level **clusters** — pupils in classrooms, patients in
clinics — and then "reliability" splits in two:

- **Subject level** (within-cluster): how reliably do raters distinguish
  *subjects within a cluster*?
- **Cluster level** (between-cluster): how reliably do raters distinguish
  *cluster means*?

Ignoring the nesting *conflates* these and biases both (ten Hove, Jorgensen & van
der Ark, 2022). Passing a `cluster` column to `icc()` fits the multilevel model
and reports each level separately.

Consider pupils nested in classrooms, each pupil rated by the same panel of
raters. We simulate a design with substantial classroom-level signal but modest
within-classroom differences:

```{r ml-data}
library(intraclass)

set.seed(2025)
n_class <- 16
n_pupil <- 5
n_rater <- 4
grid <- expand.grid(
  pupil = seq_len(n_pupil),
  classroom = seq_len(n_class),
  rater = seq_len(n_rater)
)
class_effect <- rnorm(n_class, sd = 1.3)[grid$classroom]
pupil_effect <- rnorm(n_class * n_pupil, sd = 0.6)[
  (grid$classroom - 1) * n_pupil + grid$pupil
]
rater_effect <- rnorm(n_rater, sd = 0.4)[grid$rater]
school <- data.frame(
  classroom = factor(grid$classroom),
  pupil = factor(paste(grid$classroom, grid$pupil, sep = "_")),
  rater = factor(grid$rater),
  score = 10 + class_effect + pupil_effect + rater_effect +
    rnorm(nrow(grid), sd = 0.7)
)
```

```{r ml-fit, eval = requireNamespace("glmmTMB", quietly = TRUE)}
icc(school, score, subject = pupil, rater = rater, cluster = classroom, seed = 1)
```

Both levels come back in one call. Here the **cluster-level** ICC is the higher
of the two: raters agree more about which *classrooms* score high than about which
*pupils within a classroom* do — exactly the pattern you would expect when most of
the true variation lives between classrooms. Which number you report depends on the
decision you will make: a classroom-level intervention cares about the cluster-level
reliability, a pupil-level one about the subject level. Request just one with
`level = "subject"` or `level = "cluster"`.

Multilevel support currently covers **crossed random raters on balanced data**
(agreement/consistency and single/average apply at each level). Raters nested
within clusters, incomplete multilevel designs, and fixed raters are planned for
later milestones.

## Choosing an estimation engine

By default `icc()` fits the variance components with **glmmTMB**. You can instead
request **lme4** with `engine = "lme4"` for the random two-way design. Both are
REML mixed-model fits of the same model, so on a given dataset they return the
same coefficients to numerical tolerance — the choice is about the fitting
backend, not the estimand.

```{r engine, eval = requireNamespace("lme4", quietly = TRUE) && requireNamespace("merDeriv", quietly = TRUE)}
glmmtmb <- tidy(icc(ratings, score, subject, rater, engine = "glmmTMB", seed = 1))
lme4 <- tidy(icc(ratings, score, subject, rater, engine = "lme4", seed = 1))
data.frame(
  index = glmmtmb$index,
  glmmTMB = round(glmmtmb$estimate, 4),
  lme4 = round(lme4$estimate, 4)
)
```

The two point estimates agree to well within rounding, and their Monte-Carlo
intervals coincide to about `0.01`: the lme4 interval is built from the parameter
covariance supplied by the **merDeriv** package, transformed onto the same
boundary-aware log-scale glmmTMB uses. glmmTMB remains the recommended default —
it is the one required dependency and it is robust when a variance component sits
exactly at the zero boundary, where the lme4 route cannot form an interval and
directs you back to glmmTMB. lme4 for fixed-rater and multilevel designs is
planned for a later milestone.
