---
title: "D-studies and within-cell replicates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{D-studies and within-cell replicates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(intraclass)
```

Two things you can do once you have a fitted `icc()`: project its reliability to a
different number of raters (a [**decision study**](glossary.html#d-study-decision-study)),
and — when each cell holds more than
one rating — separate the subject-by-rater interaction from pure error (**within-cell
replicates**). This article also shows the `autoplot()` methods that visualize a fit.
(Unfamiliar terms are defined in the [*Glossary*](glossary.html).)

## How many raters do I need? A D-study

`ICC(*,1)` is the reliability of a *single* rater and `ICC(*,k)` the reliability
of the mean of the `k` raters you actually used. A **decision (D-) study** asks a
forward-looking question: *how reliable would the mean of some other number of
raters `m` be?* In generalizability theory the absolute-agreement ICC is the
[dependability coefficient](glossary.html#dependability-coefficient), and projecting it
to `m` raters is just a change of the
averaging divisor,

$$\Phi(m) = \frac{\sigma^2_s}{\sigma^2_s + (\sigma^2_r + \sigma^2_{res}) / m},$$

so `d_study()` reuses the fit you already have — no refitting.

```{r dstudy, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fit <- icc(ratings, score, subject, rater, type = "agreement", seed = 1)
proj <- d_study(fit, m = 1:8)
proj
```

Reliability climbs with more raters but with diminishing returns, and the
projection is anchored to what you observed: at `m = 4` (the number of raters in
`ratings`) `Φ(m)` is exactly the `ICC(A,k)` you would get from `icc()` directly.
For a one-off value you can also ask `icc()` for it inline with a numeric `unit`,
e.g. `unit = c("single", "average", 6)`.

We fit with `type = "agreement"` here because the dependability coefficient
\eqn{\Phi(m)} above *is* the absolute-agreement projection. A default `icc()` reports
both agreement and consistency, and `d_study()` then projects **one curve per error
definition** (adding a `type` column) — handy when you want both, but for this
walkthrough one curve keeps the picture clear.

Read as a curve, this is the classic "how many raters?" picture — plot it with
`autoplot()` (which needs **ggplot2**):

```{r dstudy-plot, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("glmmTMB", quietly = TRUE), fig.alt = "Projected reliability rising with the number of raters, with a Monte-Carlo interval band."}
library(ggplot2)
autoplot(d_study(fit, m = 1:12))
```

**Projection is extrapolation.** The rater variance $\sigma^2_r$ is estimated from
only as many raters as you observed, so projecting far beyond that design leans
hard on that estimate. The Monte-Carlo interval widens honestly to reflect this
rather than pretending to a precision it lacks — and projecting absolute agreement
is refused for [*fixed* raters](glossary.html#fixed-vs.-random-raters), where there is
no wider rater universe to
generalize to (use `raters = "random"`). The [D-study also works on a multilevel
fit](multilevel-designs.html#how-many-raters-a-multilevel-d-study), projecting the
rater count at each level.

## Within-cell replicates: interaction vs. pure error

So far every subject-by-rater cell holds a single rating. When each rater rates each
subject **more than once** — *within-cell replicates* — you can separate two things
that a single rating confounds: the **subject-by-rater interaction** (does a rater
systematically score a particular subject high or low — a *stable* disagreement?)
and **pure error** (how much a rater's repeat ratings of the same subject wobble).
`icc()` detects the replicates and fits the interaction model automatically:

```{r replicates-data, eval = requireNamespace("glmmTMB", quietly = TRUE)}
set.seed(2025)
ns <- 20
nr <- 4
no <- 3
grid <- expand.grid(subject = seq_len(ns), rater = seq_len(nr), occ = seq_len(no))
subj <- rnorm(ns, sd = 1.1)[grid$subject]
rater <- rnorm(nr, sd = 0.8)[grid$rater]
sr <- rnorm(ns * nr, sd = 0.6)[(grid$rater - 1) * ns + grid$subject]
reps <- data.frame(
  subject = factor(grid$subject),
  rater = factor(grid$rater),
  score = 10 + subj + rater + sr + rnorm(nrow(grid), sd = 0.7)
)

icc(reps, score, subject, rater, type = "agreement", occasions = c("single", "average"))
```

The variance-components line now shows `subject:rater` (the interaction) and
`residual` (pure error) as separate terms. The **single-occasion** rows
(`occasions = 1`) are the ordinary ICCs — a single rating's error still includes the
interaction — but they are now fit correctly rather than folding the interaction
into the residual. The **occasion-averaged** rows (`occasions = 3` here) give the
reliability of a rater's *mean of three ratings*: averaging cuts pure error (but not
the interaction), so those coefficients are higher.

Within-cell replicates extend beyond this balanced two-way random example: **fixed**
raters (balanced), **multilevel** designs (crossed Design 1 and nested Design 2,
balanced), and **ragged** replicates at a single occasion are all supported. What
remains open is the occasion-*averaged* coefficient on ragged replicates — with unequal
per-cell counts there is no single effective-occasion divisor with a validated oracle —
along with the compound fixed-by-ragged and multilevel-by-ragged corners.

## How many occasions do I need? A D-study on the occasion facet

Just as `d_study(m = ...)` projects the number of *raters*, `d_study(n_o = ...)`
projects the number of *occasions* off a replicate fit — holding the raters fixed and
asking "how reliable would each rater's mean of `n_o` ratings be?". Supply exactly one
axis per call (`m` **or** `n_o`).

```{r occasion-dstudy, eval = requireNamespace("glmmTMB", quietly = TRUE)}
fit_rep <- icc(reps, score, subject, rater, type = "agreement", occasions = "average")
d_study(fit_rep, n_o = 1:6)
```

Notice the curve **flattens**. Averaging more occasions only cancels *pure error*
(`residual`); it never touches the rater or `subject:rater` variance. So the occasion
curve climbs to a **ceiling below 1** — the reliability you would reach with perfectly
repeatable ratings but the same raters — rather than approaching 1 the way a rater
projection does. Read it as "how much does re-rating help?", which saturates.

Because occasions are a **random** facet however the raters are treated, the occasion
projection is defined even where a rater projection is not: **fixed-rater absolute
agreement projects on the occasion axis** (it is only the *rater* axis that is
undefined for fixed absolute agreement, having no "freshly sampled rater" to add). On a
**multilevel** replicate fit the subject-level curve rises with `n_o` while the
cluster-level curve is **flat** — the cluster-level error set has no pure-error term, so
occasions cannot change it (`d_study()` says so with a note).

## Visualizing a fit

Every `icc()` fit carries an `autoplot()` method (with a `plot()` wrapper), so you
can *see* the coefficients and the variance components behind them without building
a plot by hand. Both read straight off the fitted object, so the picture can never
disagree with the printed table. They need **ggplot2**, an optional dependency.

The default, `what = "coefficients"`, is a **forest plot** — each ICC index as a
point estimate with its Monte-Carlo interval. Reusing the two-way `ratings` fit from
the D-study section above:

```{r plot-coef, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("glmmTMB", quietly = TRUE), fig.alt = "Forest plot of ICC(A,1) and ICC(A,k) for the ratings data, each a point estimate with a horizontal Monte-Carlo interval, ICC(A,k) both higher and tighter."}
library(ggplot2)
autoplot(fit) # `fit <- icc(ratings, score, subject, rater, type = "agreement", seed = 1)`
```

`what = "components"` shows the other half of the story — the estimated **variance
components** the ratio is built from. It makes plain *why* absolute agreement is so
much lower than the averaged coefficient on `ratings`: the **rater** component is
large, and only absolute agreement counts between-rater differences as error:

```{r plot-comp, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("glmmTMB", quietly = TRUE), fig.alt = "Bar chart of the estimated variance components for the ratings fit: subject, rater, and residual, with the rater component the largest."}
autoplot(fit, what = "components")
```

For a **multilevel** fit the forest plot facets by level — see [*Multilevel
designs*](multilevel-designs.html#visualizing-the-levels) for that example.
