---
title: "Estimation engines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation engines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(intraclass)
```

`icc()` fits the variance components with a mixed model by default, but the `engine`
argument lets you choose the backend. Some engine choices are purely computational —
the same estimator, a different solver — while others compute a genuinely different
(though asymptotically equivalent) estimator. This article covers when the distinction
matters.

## The mixed-model engines: glmmTMB and lme4

By default `icc()` fits the variance components with **glmmTMB**. You can instead
request **lme4** with `engine = "lme4"` for the random two-way design. Both are
REML mixed-model fits of the same model, so on a given dataset they return the
same coefficients to numerical tolerance — the choice is about the fitting
backend, not the estimand.

```{r engine, eval = requireNamespace("lme4", quietly = TRUE) && requireNamespace("merDeriv", quietly = TRUE)}
glmmtmb <- tidy(icc(ratings, score, subject, rater, engine = "glmmTMB", seed = 1))
lme4 <- tidy(icc(ratings, score, subject, rater, engine = "lme4", seed = 1))
data.frame(
  index = glmmtmb$index,
  glmmTMB = round(glmmtmb$estimate, 4),
  lme4 = round(lme4$estimate, 4)
)
```

The two point estimates agree to well within rounding, and their Monte-Carlo
intervals coincide to about `0.01`: the lme4 interval is built from the parameter
covariance supplied by the **merDeriv** package, transformed onto the same
boundary-aware log-scale glmmTMB uses. glmmTMB remains the recommended default —
it is the one required dependency and it is robust when a variance component sits
exactly at the zero boundary, where the lme4 route cannot form an interval and
directs you back to glmmTMB. lme4 otherwise has full design parity with glmmTMB — the
fixed-rater and every multilevel design, on balanced **and** incomplete/ragged data —
degrading to glmmTMB only at that variance boundary.

## A structural-equation engine (`lavaan`)

`engine = "lavaan"` fits the same design as a **structural equation model** — a
common-factor generalizability model in the sense of Jorgensen (2021) — for the
random two-way design. Unlike lme4, this is not just a different backend for the
*same* estimator: it matters which coefficient you ask for.

```{r lavaan, eval = requireNamespace("lavaan", quietly = TRUE)}
axes <- expand.grid(
  type = c("agreement", "consistency"),
  unit = c("single", "average"),
  stringsAsFactors = FALSE
)
compare <- do.call(rbind, Map(function(type, unit) {
  g <- icc(ratings, score, subject, rater, type = type, unit = unit,
           engine = "glmmTMB", seed = 1)
  l <- icc(ratings, score, subject, rater, type = type, unit = unit,
           engine = "lavaan", seed = 1)
  data.frame(
    index = tidy(g)$index,
    glmmTMB = round(tidy(g)$estimate, 4),
    lavaan = round(tidy(l)$estimate, 4)
  )
}, axes$type, axes$unit))
compare[!duplicated(compare$index), ]
```

**Consistency** coefficients are a ratio of the subject and residual variances, so
the SEM returns them identically to the mixed model. **Absolute agreement** is
different. The SEM has no random rater effect to estimate — a rater is a single
column, so its effect lives in the column *means*. Following Jorgensen (2021), the
rater variance is recovered from the mean structure as the variance of the estimated
indicator intercepts. This **indicator-mean estimator** is a genuinely different
estimator of the rater variance than the mixed model's random effect: the two are
asymptotically equivalent and match conventional generalizability-theory software
(GENOVA, `gtheory`) closely on real data [Vispoel et al. 2022], but on a small
design they differ by a modest amount — here `ICC(A,1)` is about `0.284` from lavaan
versus `0.290` from the mixed model, because the raw variance of only four estimated
rater means carries small-sample noise the mixed model shrinks away.

Which is "right"? Neither is wrong — they are two defensible estimators of the same
population quantity. Use `"glmmTMB"` (the default) if you want the mixed-model
random-rater estimate and its wider, generalize-to-new-raters interval; reach for
`"lavaan"` if you are working inside an SEM generalizability-theory workflow and want
results comparable to that literature. The SEM engine covers the random **and
fixed-rater** two-way design, on complete **and incomplete** (FIML) data — the
parametric bootstrap is available on complete data, Monte-Carlo throughout. One-way and
multilevel designs are still directed to the mixed-model engines.
