---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(intraclass)
```

## What an ICC tells you

Suppose several raters each score the same set of things --- essays, patients,
video clips --- and you want to know whether the scores can be *trusted*. If two
raters watch the same clip, will they land on the same number? If you swapped in a
different rater, would the result hold up? That is a question about **interrater
reliability**, and an **intraclass correlation coefficient (ICC)** is the number
that answers it.

An ICC runs from 0 to 1 and reports **the share of the variation in scores that
reflects real differences between the things being rated**, rather than
disagreement or noise between raters. Near 1, almost all the spread in scores is
genuine subject-to-subject difference and the raters barely disagree --- the
ratings are highly reliable. Near 0, the raters are effectively adding noise, and a
score tells you more about who happened to rate it than about the subject.

`intraclass` estimates that number by fitting a **mixed model** (it separates
subject variation from rater variation as [variance
components](glossary.html#variance-component)), rather than from
the classical ANOVA mean-squares formulas older tools use. The two agree on clean,
balanced data but the model-based approach also handles missing ratings, multiple
designs, and honest confidence intervals --- see
[*Engines*](engines.html) for the machinery. This article walks the whole pipeline
on a small example: **fit → estimate → interpret**. (New to any of the terms as they
come up? The [*Glossary*](glossary.html) defines each one in a sentence.)

## The data

We use the classic Shrout & Fleiss (1979) example, shipped with the package as
`ratings`: 6 subjects each rated by the same 4 raters, one rating per cell.
`intraclass` wants **long** format --- one rating per row, with columns for the
subject, the rater, and the score.

```{r data}
head(ratings)
```

## Fit

Call `icc()` with the data and the three columns (unquoted). The defaults give the
**two-way random, absolute-agreement** coefficients `ICC(A,1)` and `ICC(A,k)` (the
next section unpacks that label). We set a `seed` so the confidence interval is
reproducible.

```{r fit}
fit <- icc(ratings, score, subject, rater, seed = 2024)
fit
```

```{r values, include = FALSE}
td <- tidy(fit)
a1 <- td$estimate[1] # ICC(A,1), single rater
ak <- td$estimate[2] # ICC(A,k), mean of the 4 raters
ak_lo <- td$conf.low[2]
ak_hi <- td$conf.high[2]
cons1 <- tidy(icc(ratings, score, subject, rater,
  type = "consistency", seed = 2024))$estimate[1]
fmt <- function(x) sprintf("%.2f", x)
```

## Read the result with the tidy verbs

`tidy()` returns one row per coefficient, with the estimate and confidence
interval; `glance()` returns a one-row model summary including the variance
components.

```{r tidy}
tidy(fit)

glance(fit)
```

## Interpret

**Single vs. the average of several raters.** `ICC(A,1)` = `r fmt(a1)` is the
reliability of a *single* rater --- how much you can trust one person's score.
`ICC(A,k)` = `r fmt(ak)` is the reliability of the *mean* of all
`r nlevels(factor(ratings$rater))` raters. Averaging cancels out independent rater
noise, so the mean is always more reliable than one rater alone (this is the
**Spearman--Brown** relationship: more raters, higher reliability, with diminishing
returns). Report `ICC(A,k)` when the averaged score is what you will actually use,
and `ICC(A,1)` when downstream users see one rater's judgment.

**Absolute agreement vs. rank order.** These are **absolute-agreement**
coefficients: if one rater scores consistently higher than another, that systematic
gap counts as error. Here the raters differ sharply in average level, which is why
the agreement ICC is low --- a signal that the rating procedure has a level problem
worth fixing. (If you only care that raters *rank* subjects the same way, ask for
*consistency* instead --- see below.)

### Is this a good ICC?

There is no universal cutoff, but two widely cited rules of thumb give a rough
vocabulary. **Koo & Li (2016)** propose:

| ICC        | Label      |
|------------|------------|
| < 0.50     | poor       |
| 0.50--0.75 | moderate   |
| 0.75--0.90 | good       |
| > 0.90     | excellent  |

An older scheme, **Cicchetti (1994)**, draws its lines a little differently
(< 0.40 poor, 0.40--0.59 fair, 0.60--0.74 good, 0.75--1.00 excellent). Treat these
as conventions, not laws: the bar that matters depends on the stakes of your
decision, and on *which* ICC you are reading (an agreement and a consistency
coefficient on the same data are not comparable to the same cutoff).

Most important --- and this is Koo & Li's own recommendation --- **judge the
confidence interval, not just the point estimate.** Here the four-rater mean
`ICC(A,k)` = `r fmt(ak)` reads as "moderate," but its 95% interval runs from
`r fmt(ak_lo)` to `r fmt(ak_hi)` --- from "poor" all the way to "excellent." With
only six subjects, the data simply cannot pin the reliability down to one band.
That is why `icc()` never returns a point estimate without an interval, and never
labels a result for you: the honest summary is the whole interval.

### About the confidence interval

The interval `icc()` reports is a [**Monte-Carlo**
interval](glossary.html#monte-carlo-interval), and it is
[**boundary-aware**](glossary.html#zero-variance-boundary). In plain terms: instead
of relying on a textbook formula that
misbehaves when a variance is near zero (a common situation --- raters who barely
differ push the rater variance to its zero boundary), `icc()` simulates many
plausible parameter sets from the fitted model and reads the interval off the
resulting spread of ICCs. This keeps the interval well-behaved right at that
boundary. The [*Interval methods*](interval-methods.html) article covers how it
works and the alternatives (a parametric bootstrap, and Bayesian credible
intervals).

## Consistency instead of agreement

If a constant per-rater offset is acceptable --- you care only that raters *rank*
subjects the same way, not that they land on the same number --- ask for
**consistency** with `type = "consistency"`. It drops the rater main effect from
the error, so it is never smaller than the agreement coefficient:

```{r consistency}
icc(ratings, score, subject, rater, type = "consistency", seed = 2024)
```

The gap between the two (here `ICC(C,1)` = `r fmt(cons1)` vs. `ICC(A,1)` =
`r fmt(a1)`) is a direct read-out of how much systematic rater-level difference is
present.

By default raters are treated as **random** --- a sample you want to generalize
beyond, so that your reliability claim covers raters you did not use. If your raters
are the entire population of interest you can pass `raters = "fixed"` (the classic
`ICC(3,1)`), but `icc()` will warn: random is the recommended default for interrater
reliability, and on balanced data the number is identical anyway.

## Which ICC do I want?

Whether the raters are crossed (the same set judges everyone) or interchangeable
(`model = "oneway"`), absolute agreement vs. consistency, single vs. average,
fixed vs. random raters, complete vs. incomplete designs --- these choices define
*which* ICC is correct for your study. The
[*Choosing an ICC*](choosing-an-icc.html) article (`vignette("choosing-an-icc")`)
walks that decision step by step.

## References

Cicchetti, D. V. (1994). Guidelines, criteria, and rules of thumb for evaluating
normed and standardized assessment instruments in psychology. *Psychological
Assessment, 6*(4), 284--290.

Koo, T. K., & Li, M. Y. (2016). A guideline of selecting and reporting intraclass
correlation coefficients for reliability research. *Journal of Chiropractic
Medicine, 15*(2), 155--163.

Shrout, P. E., & Fleiss, J. L. (1979). Intraclass correlations: Uses in assessing
rater reliability. *Psychological Bulletin, 86*(2), 420--428.
