---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(intraclass)
```

`intraclass` estimates interrater-reliability intraclass correlation coefficients
(ICCs) from a mixed model rather than from classical ANOVA mean squares. This
article walks the full pipeline on a small, balanced example: **fit → ICC →
interpret**.

## The data

We use the classic Shrout & Fleiss (1979) example, shipped with the package as
`ratings`: 6 subjects each rated by the same 4 raters, one rating per cell.
`intraclass` wants **long** format --- one rating per row, with columns for the
subject, the rater, and the score.

```{r data}
head(ratings)
```

## Fit

Call `icc()` with the data and the three columns (unquoted). The defaults give the
**two-way random, absolute-agreement** coefficients `ICC(A,1)` and `ICC(A,k)`. We
set a `seed` so the Monte-Carlo confidence interval is reproducible.

```{r fit}
fit <- icc(ratings, score, subject, rater, seed = 2024)
fit
```

## Read the result with the tidy verbs

`tidy()` returns one row per coefficient, with the estimate and confidence
interval; `glance()` returns a one-row model summary including the variance
components.

```{r tidy}
tidy(fit)

glance(fit)
```

## Interpret

* **`ICC(A,1)` ≈ 0.29** is the reliability of a *single* rater; **`ICC(A,k)` ≈
  0.62** is the reliability of the *mean* of the 4 raters. Averaging raises
  reliability (Spearman--Brown), so report `ICC(A,k)` when the averaged score is
  what you will actually use.
* These are **absolute-agreement** coefficients: systematic differences in rater
  level (some raters scoring higher than others) count as error. Here the raters
  differ sharply in average level, which is why the agreement ICC is much lower
  than a consistency ICC would be --- a signal that the rating procedure has a
  level problem worth fixing.
* The **confidence interval** is Monte-Carlo and boundary-aware: it is simulated
  from the fitted parameter covariance on the model's internal scale, so it
  behaves near the common zero-rater-variance boundary where the delta method
  fails.

## Consistency instead of agreement

If a constant per-rater offset is acceptable --- you care only that raters *rank*
subjects the same way, not that they land on the same number --- ask for
**consistency** with `type = "consistency"`. It drops the rater main effect from
the error, so it is never smaller than the agreement coefficient:

```{r consistency}
icc(ratings, score, subject, rater, type = "consistency", seed = 2024)
```

The gap between the two (here `ICC(C,1)` ≈ 0.72 vs. `ICC(A,1)` ≈ 0.29) is a direct
read-out of how much systematic rater-level difference is present.

By default raters are treated as **random** (a sample you generalize beyond). If
your raters are the entire population of interest you can pass `raters = "fixed"`
(the classic `ICC(3,1)`), but `icc()` will warn: random is the recommended default
for interrater reliability, and on balanced data the number is identical anyway.

## Which ICC do I want?

Absolute agreement vs. consistency, single vs. average, fixed vs. random raters,
complete vs. incomplete designs --- these choices define *which* ICC is correct
for your study. The [*Choosing an ICC*](choosing-an-icc.html) article
(`vignette("choosing-an-icc")`) covers that decision in depth.
