---
title: "Confidence-interval methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidence-interval methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(intraclass)
```

Every `icc()` coefficient comes with an interval, never a bare point estimate. The
`ci_method` argument selects how that interval is built. This article covers the two
frequentist methods — Monte-Carlo and the parametric bootstrap — and the Bayesian
**credible** interval that comes with the brms engine, and when they diverge.

## Monte-Carlo and the parametric bootstrap

Every interval elsewhere in these articles has been the default **Monte-Carlo**
interval: it draws from the fitted parameter covariance on the engine's log scale and
back-transforms, which is fast and boundary-aware. A second method, a **parametric
bootstrap** (`ci_method = "bootstrap"`), instead simulates response vectors from the
fitted model, refits, and takes percentile quantiles of the resampled coefficients. It
does not lean on the asymptotic-normal covariance approximation — at the cost of a full
refit per resample, so it is far slower.

```{r ci-bootstrap, eval = requireNamespace("glmmTMB", quietly = TRUE)}
mc <- tidy(icc(ratings, score, subject, rater, seed = 1))
bs <- tidy(icc(ratings, score, subject, rater,
  ci_method = "bootstrap", boot_samples = 999, seed = 1
))
data.frame(
  index = mc$index,
  estimate = round(mc$estimate, 3),
  mc = sprintf("[%.2f, %.2f]", mc$conf.low, mc$conf.high),
  bootstrap = sprintf("[%.2f, %.2f]", bs$conf.low, bs$conf.high)
)
```

The point estimates are identical (same fit) and the upper bounds coincide; the
bootstrap's lower bounds run a little lower here, because this is a very small design
(six subjects) and the bootstrap's lower tail is noisier than the covariance-based
Monte-Carlo draw. The two methods can diverge more where the asymptotics are strained
— near the zero-variance boundary, and for the multilevel designs (more variance
components, often few clusters), where the bootstrap's cluster-level interval in
particular carries more resampling noise. The bootstrap is available for every design
the `"glmmTMB"` and `"lme4"` engines fit; `"lavaan"` supports Monte-Carlo only. Raise
`boot_samples` (default `999`) for a smoother interval at proportionally more cost.

## Bayesian credible intervals (`ci_method = "posterior"`)

When the fit is Bayesian (`engine = "brms"`, see [*Estimation
engines*](engines.html#a-bayesian-engine-brms)), the interval is neither a Monte-Carlo
nor a bootstrap *confidence* interval but a **credible** interval read directly off the
posterior draws of the ICC. `ci_method = "posterior"` is automatic — and required — for
that engine.

As in the engines article, the brms chunks below are shown with pre-computed output
(fitting a Stan model needs a toolchain not available when this site is built), so they
are not evaluated at knit time.

```{r posterior, eval = FALSE}
icc(ratings, score, subject, rater, engine = "brms", seed = 1)
```

```
#> # Intraclass correlation: two-way random, absolute agreement
#> Subjects: 6 | Raters: 4 (random) | Observations: 24 of 24 cells (complete)
#> Engine: brms (MCMC) | CI: 95% posterior credible (4000 draws)
#>   index     estimate   95% CI
#>   ICC(A,1)    0.241   [0.066, 0.649]
#>   ICC(A,k)    0.679   [0.221, 0.881]
#> Variance components: subject 1.522, rater 2.653, residual 0.962
#> Shrout & Fleiss equivalent: ICC(A,1) = ICC(2,1), ICC(A,k) = ICC(2,k)
```

The point estimate is the posterior mode (MAP) and the default interval is a
**percentile** credible interval — the lower `2.5%` and upper `97.5%` quantiles of the
ICC draws.
Percentile is the default because it is invariant to how the ICC is parameterized and
degrades gracefully as a variance component approaches zero (ten Hove et al. 2020, §4.2,
find it nominal at more than two raters).

### Highest-posterior-density intervals

For comparison you can ask for a **highest-posterior-density interval** (HPDI) — the
*narrowest* interval containing 95% of the posterior mass — with
`posterior_summary = "hpdi"`:

```{r posterior-hpdi, eval = FALSE}
icc(ratings, score, subject, rater, engine = "brms",
  posterior_summary = "hpdi", seed = 1)
```

```
#> # Intraclass correlation: two-way random, absolute agreement
#> Subjects: 6 | Raters: 4 (random) | Observations: 24 of 24 cells (complete)
#> Engine: brms (MCMC) | CI: 95% posterior credible (HPDI) (4000 draws)
#>   index     estimate   95% CI
#>   ICC(A,1)    0.241   [0.040, 0.601]
#>   ICC(A,k)    0.679   [0.256, 0.904]
#> Variance components: subject 1.522, rater 2.653, residual 0.962
#> Shrout & Fleiss equivalent: ICC(A,1) = ICC(2,1), ICC(A,k) = ICC(2,k)
```

The header now flags `(HPDI)`, and on the same draws the interval is no wider than the
percentile one — that is what "narrowest" means. Here `ICC(A,1)` is `[0.04, 0.60]`
against the percentile `[0.07, 0.65]`; the point estimate (the MAP) is unchanged.
Percentile stays the default, because HPDI is not transform-invariant and can behave
less well at the variance boundary — it is offered for comparison, not as an upgrade.
