---
title: "Confidence-interval methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidence-interval methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(intraclass)
```

A point estimate on its own can mislead — with a handful of subjects the same ICC
could be "poor" or "excellent" and you would not know it. So every `icc()` coefficient
comes with an interval, never a bare number, and the `ci_method` argument selects how
that interval is built. This article covers the Monte-Carlo default and the
parametric bootstrap; the [four opt-in
methods](#the-opt-in-boundary-robust-methods) for the near-boundary terrain the
default struggles on — the transformed bootstrap-*t* (`"npbootstrap"`), the exact-F
(`"searle"`) and Burch (`"burch"`) closed forms, and the modified profile likelihood
(`"mpl"`); and the Bayesian [**credible**](glossary.html#credible-interval) interval
that comes with the brms engine. (Terms are defined in the
[*Glossary*](glossary.html).)

## Monte-Carlo and the parametric bootstrap

Every interval elsewhere in these articles has been the default [**Monte-Carlo**
interval](glossary.html#monte-carlo-interval): it draws from the fitted parameter
covariance on the engine's log scale and
back-transforms, which is fast and [boundary-aware](glossary.html#zero-variance-boundary).
A second method, a [**parametric
bootstrap**](glossary.html#parametric-bootstrap) (`ci_method = "bootstrap"`), instead
simulates response vectors from the
fitted model, refits, and takes percentile quantiles of the resampled coefficients. It
does not lean on the asymptotic-normal covariance approximation (the assumption that the
estimates are normally distributed around the truth, which frays in small samples) — at
the cost of a full refit per resample, so it is far slower.

```{r ci-bootstrap, eval = requireNamespace("glmmTMB", quietly = TRUE)}
mc <- tidy(icc(ratings, score, subject, rater, seed = 1))
bs <- tidy(icc(ratings, score, subject, rater,
  ci_method = "bootstrap", boot_samples = 999, seed = 1
))
data.frame(
  index = mc$index,
  estimate = round(mc$estimate, 3),
  mc = sprintf("[%.2f, %.2f]", mc$conf.low, mc$conf.high),
  bootstrap = sprintf("[%.2f, %.2f]", bs$conf.low, bs$conf.high)
)
```

The point estimates are identical (same fit) and the upper bounds coincide; the
bootstrap's lower bounds run a little lower here, because this is a very small design
(six subjects) and the bootstrap's lower tail is noisier than the covariance-based
Monte-Carlo draw. The two methods can diverge more where the asymptotics are strained
— near the zero-variance boundary, and for the multilevel designs (more variance
components, often few clusters), where the bootstrap's cluster-level interval in
particular carries more resampling noise. The bootstrap is available for every design
the `"glmmTMB"` and `"lme4"` engines fit; `"lavaan"` supports Monte-Carlo only. Raise
`boot_samples` (default `999`) for a smoother interval at proportionally more cost.

### When the default under-covers

The boundary is not the default's only weak spot, and the other one is easier to
miss because nothing about the output looks wrong. The Monte-Carlo draw assumes the
fitted parameters are approximately normally distributed around the truth. That
assumption is about the *estimates*, not the ratings — but it degrades when the
**subject effects** are strongly skewed or heavy-tailed, and then a nominal 95%
interval covers considerably less often than 95% of the time.

A one-way simulation study measured this across four subject-effect distributions.
Where the default produced an interval at all, coverage fell to 0.6725 at its worst
— chi-square(1) subject effects, a true ICC of 0.6, 50 subjects and 5 raters. Those
runs did not abort, warn, or widen, so the shortfall is invisible in the interval
itself.

Two patterns in that study are worth carrying away, and one tempting reading of it
is wrong. At 5 raters per subject, coverage falls as the subject count rises, once
the true ICC is moderate or high — the largest designs are the exposed ones, not the
smallest. And near-normal or uniform subject effects under-covered only in cells
where many runs aborted; wherever the default almost always returned an interval,
those distributions were fine. The wrong reading is that fewer raters is safer: in
every cell where both were measured, 2 raters covered worse than 5. What changes is
that a larger share of the 2-rater runs abort outright, and an abort is a visible
failure rather than a quiet one.

The held-out battery agrees where the geometry matches: lognormal and Laplace
subject effects covered 0.825 and 0.84 at that same 50-subject, 5-rater geometry,
while their 20-subject, 3-rater cells came out near nominal.

The natural reaction — switch to a closed form — does not help. In every cell where
the default under-covered without also aborting often, `"searle"` and `"burch"`
under-covered as well, usually by more: `"burch"`, the one this article used to
recommend for heavy tails, bottoms out at 0.6655. The remaining methods were never
run on that study, so this article recommends none of them in its place.

What to do instead is ordinary statistical hygiene rather than a package setting.
Look at the distribution of the subject means before trusting a narrow interval;
report the variance components alongside the coefficient; and treat an interval on
visibly skewed data as optimistic about its own precision.

## The opt-in boundary-robust methods

Near the [zero-variance boundary](glossary.html#zero-variance-boundary) the
Monte-Carlo default can fail to produce an interval; when it aborts, its message
names an alternative method — chosen by running the candidates on your own data,
not by consulting a table — so in practice you rarely need to pick one from
scratch. This section is for when you do. Four opt-in methods serve exactly that
terrain; each is fenced to a specific design and aborts with a classed error
anywhere else.

### The transformed bootstrap-*t* (`ci_method = "npbootstrap"`)

The non-parametric [**transformed
bootstrap-*t***](glossary.html#transformed-bootstrap-t) of Ukoumunne et al. (2003)
serves the **one-way random design** (`model = "oneway"`), on balanced and
unbalanced data alike — of the four opt-in methods it is the only one that
serves unbalanced one-way data. It resamples whole subjects with replacement rather than
simulating from the fitted model, so it is the only opt-in method that takes a
`seed` (and `boot_samples`); pin both for a reproducible interval. Any
`conf_level` in `(0, 1)` is accepted. `unit = "average"` — the ICC(k) — is the
exact monotone Spearman-Brown image of the ICC(1) endpoints, so its coverage is
inherited by construction, balanced or not; a numeric `unit` (a D-study
projection) is restricted to balanced data. Reach for it for boundary robustness
— an interval that exists where the Monte-Carlo default aborts — and for
robustness to non-normal subject effects. Its endpoints are deliberately left
untruncated on the estimator's own support (Ukoumunne et al. 2003, §5.2), so a
near-boundary lower limit can be negative — honest disclosure, not an error.

### The classical closed forms (`ci_method = "searle"` and `"burch"`)

Two deterministic classical intervals for the **balanced one-way random design**:
closed forms with no resampling, so `mc_samples`, `boot_samples`, and `seed` do
not apply and no `std.error` is reported. Any `conf_level` in `(0, 1)` is
accepted, and both project ICC(k) — and a numeric `unit` — through the same
Spearman-Brown image as `"npbootstrap"`. The [**exact-F
interval**](glossary.html#exact-f-interval) (`"searle"`; Searle 1971, the McGraw &
Wong 1996 Table 7 limits) is exact under normality and best-calibrated when the
data are approximately normal. The [**Burch
interval**](glossary.html#burch-interval) (`"burch"`; Burch 2011) is REML-based
and kurtosis-adjusted: its width tracks the data's tail weight, which buys it
some robustness to mild non-normality. It is not, however, a remedy for heavy
tails: on strongly skewed subject effects it under-covers about as badly as the
default (see [When the default under-covers](#when-the-default-under-covers)).

**Which is the tighter interval?** Neither, reliably — and the margin between
them is not a fixed one. Over the larger grid's 64 cells, spanning four
distribution families, `"burch"` is the *narrower* of the two in 59 of 64 cells
of the larger grid, but how much narrower depends on where in the design you
look. Below, the median width ratio is `"burch"`'s width over `"searle"`'s, so a
value under 1 means `"burch"` is narrower and 1 means they are the same width.

| true ICC | median width ratio | `"burch"` narrower |
|---:|---:|---:|
| 0.05 | 0.9485 | 16 of 16 |
| 0.1 | 0.9470 | 16 of 16 |
| 0.3 | 0.9475 | 16 of 16 |
| 0.6 | 0.9971 | 11 of 16 |

| subjects (at 5 raters) | median width ratio | `"burch"` narrower |
|---:|---:|---:|
| 10 | 0.9154 | 15 of 16 |
| 30 | 0.9646 | 15 of 16 |
| 50 | 0.9769 | 13 of 16 |

Two things to read off. In the true ICC, `"burch"`'s width advantage holds much
the same up to a true ICC of 0.3 rather than shrinking as the true ICC rises —
the largest margin is at a true ICC of 0.1, not at the bottom of the range — and
then collapses to near parity at a true ICC of 0.6, on the one grid reaching that
value, where every cell favouring `"searle"` sits. In the subject count it
shrinks steadily as the subject count grows, measured at 5 raters, which is the
cut the second table takes. That cut is at 5 raters because 10 subjects is the
only subject count in either grid where the rater count varies, so an
unstratified row there would be confounded with it. The smaller grid's 16 cells
carry only the two lowest true-ICC values, so that sweep cannot show the first
pattern; it shows the second in the same direction, `"burch"` running narrower by
a median width ratio of 0.9017 at 10 subjects, 0.9611 at 30 subjects and 0.9775
at 50 subjects.

A pooled figure over both grids would hide all of this, and it would invite a
second misreading. The smaller grid's design points are a subset of the larger
one's, so much of the gap between their pooled medians is which true-ICC values
each grid covers rather than a disagreement between two bodies of evidence — but
only much of it. Restricting the larger grid to the smaller one's design points
closes most of that gap and leaves a remainder, and the two are separate
simulations that mostly disagree at the design points they share, agreeing
closely at only a couple of them. A pooled between-grid comparison is not a clean
contrast in either direction, which is why nothing above is one.

There is a real limit on the comparison, though, because both
grids draw only the **subject effects** from the non-normal family and always
draw the errors from a normal. Burch's own expected-length comparison — which is
against this very exact-F interval — is kurtosis-conditional: he finds his
interval shorter for light-tailed data but *wider* for symmetric heavy-tailed
data, measuring with both the subject effects and the errors non-normal. So the
honest summary is that the ordering depends on the data, this package has
measured it only in the subject-effect-only case, and you should not pick between
them on width.

Prefer `"searle"`: across every distribution family in that skew study it landed
closer to nominal coverage in most cells, heavy-tailed ones included. What
`"burch"` buys is dipping below the nominal level in fewer cells overall — a more
limited kind of steadiness than its kurtosis adjustment suggests. Their value over
the default is a finite, well-calibrated interval at the near-zero-ICC boundary
where the Monte-Carlo default aborts. One asymmetry
between the siblings: on data with *no* between-subject variance at all,
`"burch"` aborts (its kurtosis standardization divides by zero there) while
`"searle"` still returns its attained minimum.

```{r ci-oneway-optin, eval = requireNamespace("glmmTMB", quietly = TRUE)}
mc <- tidy(icc(ratings, score, subject, rater, model = "oneway", seed = 1))
se <- tidy(icc(ratings, score, subject, rater,
  model = "oneway", ci_method = "searle"
))
bu <- tidy(icc(ratings, score, subject, rater,
  model = "oneway", ci_method = "burch"
))
np <- tidy(icc(ratings, score, subject, rater,
  model = "oneway", ci_method = "npbootstrap", boot_samples = 199, seed = 1
))
data.frame(
  index = mc$index,
  estimate = round(mc$estimate, 3),
  montecarlo = sprintf("[%.2f, %.2f]", mc$conf.low, mc$conf.high),
  searle = sprintf("[%.2f, %.2f]", se$conf.low, se$conf.high),
  burch = sprintf("[%.2f, %.2f]", bu$conf.low, bu$conf.high),
  npbootstrap = sprintf("[%.2f, %.2f]", np$conf.low, np$conf.high)
)
```

All four columns share the same point estimate — `ci_method` selects the
interval, never the estimator. The visible difference is at the lower end: the
three opt-in methods' lower limits dip below zero, because their endpoints are
left untruncated on the estimator's own support, while the Monte-Carlo interval
stays inside the range. And the Burch adjustment is empirical, not a one-way
widening — here it comes out *narrower* than the exact-F interval, because its
width tracks the tail weight these particular data actually show, which is the
same direction the grids above measure.

### The modified profile likelihood (`ci_method = "mpl"`)

The [**modified profile-likelihood
interval**](glossary.html#modified-profile-likelihood) of Xiao & Liu (2013) is
the two-way counterpart: it serves the **balanced, complete two-way random
absolute-agreement** ICC(A,1), with ICC(A,k) and any numeric-`unit` projection
its pole-safe Spearman-Brown image, and aborts on any other design, on
consistency or fixed raters, and on unbalanced or incomplete data. It is a
deterministic closed form — no resampling, no `seed`. Its calibration fixes two
fences: `conf_level` must be 0.90, 0.95, or 0.99 (each level carries its own
calibrated correction constant, never interpolated between levels), and the
calibration grid spans 2–10 raters and 10–100 subjects. Like `"npbootstrap"`, it
returns an interval at the near-zero-ICC boundary where the two-way Monte-Carlo
default aborts; it is deliberately conservative — it over-covers, and is wider
than the Monte-Carlo interval at interior cells — which is why it is an opt-in
and not the default. Two reporting caveats from `?icc`: the two-sided interval
is not equal-tailed, so a limit must not be read as a one-sided bound at half
the complementary level; and at `conf_level = 0.99` with two raters the interval
can be near-vacuous.

The shipped `ratings` data are too small for the calibration grid (six
subjects), so the demonstration simulates a balanced two-way design inside it:

```{r ci-mpl, eval = requireNamespace("glmmTMB", quietly = TRUE)}
set.seed(88)
n_s <- 20
n_r <- 4
subj_eff <- rnorm(n_s, sd = sqrt(0.6))
rater_eff <- rnorm(n_r, sd = sqrt(0.1))
noise <- matrix(rnorm(n_s * n_r, sd = sqrt(0.2)), n_s, n_r)
sim <- data.frame(
  subject = factor(rep(seq_len(n_s), times = n_r)),
  rater = factor(rep(seq_len(n_r), each = n_s)),
  score = as.numeric(outer(subj_eff, rep(1, n_r)) +
    outer(rep(1, n_s), rater_eff) + noise)
)
mc2 <- tidy(icc(sim, score, subject, rater, type = "agreement", seed = 1))
ml <- tidy(icc(sim, score, subject, rater, type = "agreement", ci_method = "mpl"))
data.frame(
  index = mc2$index,
  estimate = round(mc2$estimate, 3),
  montecarlo = sprintf("[%.2f, %.2f]", mc2$conf.low, mc2$conf.high),
  mpl = sprintf("[%.2f, %.2f]", ml$conf.low, ml$conf.high)
)
```

The two point estimates agree (same REML fit); the `"mpl"` interval is the wider
of the pair at this comfortably interior cell — the conservatism described
above, visible on ordinary data.

## Bayesian credible intervals (`ci_method = "posterior"`)

When the fit is Bayesian (`engine = "brms"`, see [*Estimation
engines*](engines.html#a-bayesian-engine-brms)), the interval is neither a Monte-Carlo
nor a bootstrap *confidence* interval but a **credible** interval read directly off the
posterior draws of the ICC — a [different kind of
statement](glossary.html#confidence-interval-vs.-credible-interval) about where the ICC
lies. `ci_method = "posterior"` is automatic — and required — for
that engine.

As in the engines article, the brms chunks below are shown with pre-computed output
(fitting a Stan model needs a toolchain not available when this site is built), so they
are not evaluated at knit time.

```{r posterior, eval = FALSE}
icc(ratings, score, subject, rater, engine = "brms", type = "agreement", seed = 1)
```

```
#> ── Intraclass correlation: two-way random, absolute agreement ──────────────────
#> Subjects: 6 | Raters: 4 (random) | Observations: 24 of 24 cells (complete)
#> Engine: brms (MCMC) | CI: 95% posterior credible (4000 draws)
#>
#>   index     estimate   95% CI
#>   ICC(A,1)     0.241   [0.066, 0.649]
#>   ICC(A,k)     0.679   [0.221, 0.881]
#>
#> Variance components: subject 1.522, rater 2.653, residual 0.962
#> Shrout & Fleiss equivalent: ICC(A,1) = ICC(2,1), ICC(A,k) = ICC(2,k)
```

The point estimate is the [posterior mode (MAP)](glossary.html#posterior-mode-map) and
the default interval is a
**percentile** credible interval — the lower `2.5%` and upper `97.5%` quantiles of the
ICC draws.
Percentile is the default because it is invariant to how the ICC is parameterized and
degrades gracefully as a variance component approaches zero (ten Hove et al. 2020, §4.2,
find it nominal at more than two raters).

### Highest-posterior-density intervals

For comparison you can ask for a **highest-posterior-density interval** (HPDI) — the
*narrowest* interval containing 95% of the posterior mass — with
`posterior_summary = "hpdi"`:

```{r posterior-hpdi, eval = FALSE}
icc(ratings, score, subject, rater, engine = "brms",
  type = "agreement", posterior_summary = "hpdi", seed = 1)
```

```
#> ── Intraclass correlation: two-way random, absolute agreement ──────────────────
#> Subjects: 6 | Raters: 4 (random) | Observations: 24 of 24 cells (complete)
#> Engine: brms (MCMC) | CI: 95% posterior credible (HPDI) (4000 draws)
#>
#>   index     estimate   95% CI
#>   ICC(A,1)     0.241   [0.040, 0.601]
#>   ICC(A,k)     0.679   [0.256, 0.904]
#>
#> Variance components: subject 1.522, rater 2.653, residual 0.962
#> Shrout & Fleiss equivalent: ICC(A,1) = ICC(2,1), ICC(A,k) = ICC(2,k)
```

The header now flags `(HPDI)`, and on the same draws the interval is no wider than the
percentile one — that is what "narrowest" means. Here `ICC(A,1)` is `[0.04, 0.60]`
against the percentile `[0.07, 0.65]`; the point estimate (the MAP) is unchanged.
Percentile stays the default, because HPDI is not transform-invariant and can behave
less well at the variance boundary — it is offered for comparison, not as an upgrade.
