---
title: "Confidence-interval methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidence-interval methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(intraclass)
```

A point estimate on its own can mislead — with a handful of subjects the same ICC
could be "poor" or "excellent" and you would not know it. So every `icc()` coefficient
comes with an interval, never a bare number, and the `ci_method` argument selects how
that interval is built. This article covers the two default frequentist methods —
Monte-Carlo and the parametric bootstrap; the [four opt-in
methods](#the-opt-in-boundary-robust-methods) for the near-boundary terrain the
default struggles on — the transformed bootstrap-*t* (`"npbootstrap"`), the exact-F
(`"searle"`) and Burch (`"burch"`) closed forms, and the modified profile likelihood
(`"mpl"`); and the Bayesian [**credible**](glossary.html#credible-interval) interval
that comes with the brms engine. (Terms are defined in the
[*Glossary*](glossary.html).)

## Monte-Carlo and the parametric bootstrap

Every interval elsewhere in these articles has been the default [**Monte-Carlo**
interval](glossary.html#monte-carlo-interval): it draws from the fitted parameter
covariance on the engine's log scale and
back-transforms, which is fast and [boundary-aware](glossary.html#zero-variance-boundary).
A second method, a [**parametric
bootstrap**](glossary.html#parametric-bootstrap) (`ci_method = "bootstrap"`), instead
simulates response vectors from the
fitted model, refits, and takes percentile quantiles of the resampled coefficients. It
does not lean on the asymptotic-normal covariance approximation (the assumption that the
estimates are normally distributed around the truth, which frays in small samples) — at
the cost of a full refit per resample, so it is far slower.

```{r ci-bootstrap, eval = requireNamespace("glmmTMB", quietly = TRUE)}
mc <- tidy(icc(ratings, score, subject, rater, seed = 1))
bs <- tidy(icc(ratings, score, subject, rater,
  ci_method = "bootstrap", boot_samples = 999, seed = 1
))
data.frame(
  index = mc$index,
  estimate = round(mc$estimate, 3),
  mc = sprintf("[%.2f, %.2f]", mc$conf.low, mc$conf.high),
  bootstrap = sprintf("[%.2f, %.2f]", bs$conf.low, bs$conf.high)
)
```

The point estimates are identical (same fit) and the upper bounds coincide; the
bootstrap's lower bounds run a little lower here, because this is a very small design
(six subjects) and the bootstrap's lower tail is noisier than the covariance-based
Monte-Carlo draw. The two methods can diverge more where the asymptotics are strained
— near the zero-variance boundary, and for the multilevel designs (more variance
components, often few clusters), where the bootstrap's cluster-level interval in
particular carries more resampling noise. The bootstrap is available for every design
the `"glmmTMB"` and `"lme4"` engines fit; `"lavaan"` supports Monte-Carlo only. Raise
`boot_samples` (default `999`) for a smoother interval at proportionally more cost.

## The opt-in boundary-robust methods

Near the [zero-variance boundary](glossary.html#zero-variance-boundary) the
Monte-Carlo default can fail to produce an interval; when it aborts, its message
names an alternative method — chosen by running the candidates on your own data,
not by consulting a table — so in practice you rarely need to pick one from
scratch. This section is for when you do. Four opt-in methods serve exactly that
terrain; each is fenced to a specific design and aborts with a classed error
anywhere else.

### The transformed bootstrap-*t* (`ci_method = "npbootstrap"`)

The non-parametric [**transformed
bootstrap-*t***](glossary.html#transformed-bootstrap-t) of Ukoumunne et al. (2003)
serves the **one-way random design** (`model = "oneway"`), on balanced and
unbalanced data alike — on an unbalanced one-way design it is the one interval
method available. It resamples whole subjects with replacement rather than
simulating from the fitted model, so it is the only opt-in method that takes a
`seed` (and `boot_samples`); pin both for a reproducible interval. Any
`conf_level` in `(0, 1)` is accepted. `unit = "average"` — the ICC(k) — is the
exact monotone Spearman-Brown image of the ICC(1) endpoints, so its coverage is
inherited by construction, balanced or not; a numeric `unit` (a D-study
projection) is restricted to balanced data. Reach for it for boundary robustness
— an interval that exists where the Monte-Carlo default aborts — and for
robustness to non-normal subject effects. Its endpoints are deliberately left
untruncated on the estimator's own support (Ukoumunne et al. 2003, §5.2), so a
near-boundary lower limit can be negative — honest disclosure, not an error.

### The classical closed forms (`ci_method = "searle"` and `"burch"`)

Two deterministic classical intervals for the **balanced one-way random design**:
closed forms with no resampling, so `mc_samples`, `boot_samples`, and `seed` do
not apply and no `std.error` is reported. Any `conf_level` in `(0, 1)` is
accepted, and both project ICC(k) — and a numeric `unit` — through the same
Spearman-Brown image as `"npbootstrap"`. The [**exact-F
interval**](glossary.html#exact-f-interval) (`"searle"`; Searle 1971, the McGraw &
Wong 1996 Table 7 limits) is exact under normality: best-calibrated and narrowest
when the data are approximately normal. The [**Burch
interval**](glossary.html#burch-interval) (`"burch"`; Burch 2011) is REML-based
and kurtosis-adjusted: wider, but robust to non-normality. Prefer `"searle"` for
near-normal data and `"burch"` when heavy tails are a concern. Their value over
the default is a finite, well-calibrated interval at the near-zero-ICC boundary
where the Monte-Carlo default aborts. One asymmetry
between the siblings: on data with *no* between-subject variance at all,
`"burch"` aborts (its kurtosis standardization divides by zero there) while
`"searle"` still returns its attained minimum.

```{r ci-oneway-optin, eval = requireNamespace("glmmTMB", quietly = TRUE)}
mc <- tidy(icc(ratings, score, subject, rater, model = "oneway", seed = 1))
se <- tidy(icc(ratings, score, subject, rater,
  model = "oneway", ci_method = "searle"
))
bu <- tidy(icc(ratings, score, subject, rater,
  model = "oneway", ci_method = "burch"
))
np <- tidy(icc(ratings, score, subject, rater,
  model = "oneway", ci_method = "npbootstrap", boot_samples = 199, seed = 1
))
data.frame(
  index = mc$index,
  estimate = round(mc$estimate, 3),
  montecarlo = sprintf("[%.2f, %.2f]", mc$conf.low, mc$conf.high),
  searle = sprintf("[%.2f, %.2f]", se$conf.low, se$conf.high),
  burch = sprintf("[%.2f, %.2f]", bu$conf.low, bu$conf.high),
  npbootstrap = sprintf("[%.2f, %.2f]", np$conf.low, np$conf.high)
)
```

All four columns share the same point estimate — `ci_method` selects the
interval, never the estimator. The visible difference is at the lower end: the
three opt-in methods' lower limits dip below zero, because their endpoints are
left untruncated on the estimator's own support, while the Monte-Carlo interval
stays inside the range. And the Burch adjustment is empirical, not a one-way
widening — here it comes out *narrower* than the exact-F interval, because its
width tracks the tail weight these particular data actually show.

### The modified profile likelihood (`ci_method = "mpl"`)

The [**modified profile-likelihood
interval**](glossary.html#modified-profile-likelihood) of Xiao & Liu (2013) is
the two-way counterpart: it serves the **balanced, complete two-way random
absolute-agreement** ICC(A,1), with ICC(A,k) and any numeric-`unit` projection
its pole-safe Spearman-Brown image, and aborts on any other design, on
consistency or fixed raters, and on unbalanced or incomplete data. It is a
deterministic closed form — no resampling, no `seed`. Its calibration fixes two
fences: `conf_level` must be 0.90, 0.95, or 0.99 (each level carries its own
calibrated correction constant, never interpolated between levels), and the
calibration grid spans 2–10 raters and 10–100 subjects. Like `"npbootstrap"`, it
returns an interval at the near-zero-ICC boundary where the two-way Monte-Carlo
default aborts; it is deliberately conservative — it over-covers, and is wider
than the Monte-Carlo interval at interior cells — which is why it is an opt-in
and not the default. Two reporting caveats from `?icc`: the two-sided interval
is not equal-tailed, so a limit must not be read as a one-sided bound at half
the complementary level; and at `conf_level = 0.99` with two raters the interval
can be near-vacuous.

The shipped `ratings` data are too small for the calibration grid (six
subjects), so the demonstration simulates a balanced two-way design inside it:

```{r ci-mpl, eval = requireNamespace("glmmTMB", quietly = TRUE)}
set.seed(88)
n_s <- 20
n_r <- 4
subj_eff <- rnorm(n_s, sd = sqrt(0.6))
rater_eff <- rnorm(n_r, sd = sqrt(0.1))
noise <- matrix(rnorm(n_s * n_r, sd = sqrt(0.2)), n_s, n_r)
sim <- data.frame(
  subject = factor(rep(seq_len(n_s), times = n_r)),
  rater = factor(rep(seq_len(n_r), each = n_s)),
  score = as.numeric(outer(subj_eff, rep(1, n_r)) +
    outer(rep(1, n_s), rater_eff) + noise)
)
mc2 <- tidy(icc(sim, score, subject, rater, type = "agreement", seed = 1))
ml <- tidy(icc(sim, score, subject, rater, type = "agreement", ci_method = "mpl"))
data.frame(
  index = mc2$index,
  estimate = round(mc2$estimate, 3),
  montecarlo = sprintf("[%.2f, %.2f]", mc2$conf.low, mc2$conf.high),
  mpl = sprintf("[%.2f, %.2f]", ml$conf.low, ml$conf.high)
)
```

The two point estimates agree (same REML fit); the `"mpl"` interval is the wider
of the pair at this comfortably interior cell — the conservatism described
above, visible on ordinary data.

## Bayesian credible intervals (`ci_method = "posterior"`)

When the fit is Bayesian (`engine = "brms"`, see [*Estimation
engines*](engines.html#a-bayesian-engine-brms)), the interval is neither a Monte-Carlo
nor a bootstrap *confidence* interval but a **credible** interval read directly off the
posterior draws of the ICC — a [different kind of
statement](glossary.html#confidence-interval-vs.-credible-interval) about where the ICC
lies. `ci_method = "posterior"` is automatic — and required — for
that engine.

As in the engines article, the brms chunks below are shown with pre-computed output
(fitting a Stan model needs a toolchain not available when this site is built), so they
are not evaluated at knit time.

```{r posterior, eval = FALSE}
icc(ratings, score, subject, rater, engine = "brms", type = "agreement", seed = 1)
```

```
#> ── Intraclass correlation: two-way random, absolute agreement ──────────────────
#> Subjects: 6 | Raters: 4 (random) | Observations: 24 of 24 cells (complete)
#> Engine: brms (MCMC) | CI: 95% posterior credible (4000 draws)
#>
#>   index     estimate   95% CI
#>   ICC(A,1)     0.241   [0.066, 0.649]
#>   ICC(A,k)     0.679   [0.221, 0.881]
#>
#> Variance components: subject 1.522, rater 2.653, residual 0.962
#> Shrout & Fleiss equivalent: ICC(A,1) = ICC(2,1), ICC(A,k) = ICC(2,k)
```

The point estimate is the [posterior mode (MAP)](glossary.html#posterior-mode-map) and
the default interval is a
**percentile** credible interval — the lower `2.5%` and upper `97.5%` quantiles of the
ICC draws.
Percentile is the default because it is invariant to how the ICC is parameterized and
degrades gracefully as a variance component approaches zero (ten Hove et al. 2020, §4.2,
find it nominal at more than two raters).

### Highest-posterior-density intervals

For comparison you can ask for a **highest-posterior-density interval** (HPDI) — the
*narrowest* interval containing 95% of the posterior mass — with
`posterior_summary = "hpdi"`:

```{r posterior-hpdi, eval = FALSE}
icc(ratings, score, subject, rater, engine = "brms",
  type = "agreement", posterior_summary = "hpdi", seed = 1)
```

```
#> ── Intraclass correlation: two-way random, absolute agreement ──────────────────
#> Subjects: 6 | Raters: 4 (random) | Observations: 24 of 24 cells (complete)
#> Engine: brms (MCMC) | CI: 95% posterior credible (HPDI) (4000 draws)
#>
#>   index     estimate   95% CI
#>   ICC(A,1)     0.241   [0.040, 0.601]
#>   ICC(A,k)     0.679   [0.256, 0.904]
#>
#> Variance components: subject 1.522, rater 2.653, residual 0.962
#> Shrout & Fleiss equivalent: ICC(A,1) = ICC(2,1), ICC(A,k) = ICC(2,k)
```

The header now flags `(HPDI)`, and on the same draws the interval is no wider than the
percentile one — that is what "narrowest" means. Here `ICC(A,1)` is `[0.04, 0.60]`
against the percentile `[0.07, 0.65]`; the point estimate (the MAP) is unchanged.
Percentile stays the default, because HPDI is not transform-invariant and can behave
less well at the variance boundary — it is offered for comparison, not as an upgrade.
