---
title: "Confidence-interval methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidence-interval methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(intraclass)
```

Every `icc()` coefficient comes with an interval, never a bare point estimate. The
`ci_method` argument selects how that interval is built. This article covers the two
available methods and when they diverge.

## Monte-Carlo and the parametric bootstrap

Every interval elsewhere in these articles has been the default **Monte-Carlo**
interval: it draws from the fitted parameter covariance on the engine's log scale and
back-transforms, which is fast and boundary-aware. A second method, a **parametric
bootstrap** (`ci_method = "bootstrap"`), instead simulates response vectors from the
fitted model, refits, and takes percentile quantiles of the resampled coefficients. It
does not lean on the asymptotic-normal covariance approximation — at the cost of a full
refit per resample, so it is far slower.

```{r ci-bootstrap, eval = requireNamespace("glmmTMB", quietly = TRUE)}
mc <- tidy(icc(ratings, score, subject, rater, seed = 1))
bs <- tidy(icc(ratings, score, subject, rater,
  ci_method = "bootstrap", boot_samples = 999, seed = 1
))
data.frame(
  index = mc$index,
  estimate = round(mc$estimate, 3),
  mc = sprintf("[%.2f, %.2f]", mc$conf.low, mc$conf.high),
  bootstrap = sprintf("[%.2f, %.2f]", bs$conf.low, bs$conf.high)
)
```

The point estimates are identical (same fit) and the upper bounds coincide; the
bootstrap's lower bounds run a little lower here, because this is a very small design
(six subjects) and the bootstrap's lower tail is noisier than the covariance-based
Monte-Carlo draw. The two methods can diverge more where the asymptotics are strained
— near the zero-variance boundary, and for the multilevel designs (more variance
components, often few clusters), where the bootstrap's cluster-level interval in
particular carries more resampling noise. The bootstrap is available for every design
the `"glmmTMB"` and `"lme4"` engines fit; `"lavaan"` supports Monte-Carlo only. Raise
`boot_samples` (default `999`) for a smoother interval at proportionally more cost.
