---
title: "Multilevel designs: subject and cluster level"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multilevel designs: subject and cluster level}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(intraclass)
```

The [*Getting started*](getting-started.html) and [*Choosing an ICC*](choosing-an-icc.html)
articles treat the **subject** as the object of measurement. But subjects are often
nested in higher-level **clusters** — pupils in classrooms, patients in clinics — and
then "reliability" splits in two. This article covers the multilevel ICC family (ten
Hove, Jorgensen & van der Ark, 2022): crossed and nested layouts, complete and
incomplete data, fixed raters, and the multilevel D-study.

## Subject level vs. cluster level

When subjects are nested in clusters, two reliabilities are defined:

- **Subject level** (within-cluster): how reliably do raters distinguish
  *subjects within a cluster*?
- **Cluster level** (between-cluster): how reliably do raters distinguish
  *cluster means*?

Ignoring the nesting *conflates* these and biases both (ten Hove, Jorgensen & van
der Ark, 2022). Passing a `cluster` column to `icc()` fits the multilevel model
and reports each level separately.

Consider pupils nested in classrooms, each pupil rated by the same panel of
raters. We simulate a design with substantial classroom-level signal but modest
within-classroom differences:

```{r ml-data}
set.seed(2025)
n_class <- 16
n_pupil <- 5
n_rater <- 4
grid <- expand.grid(
  pupil = seq_len(n_pupil),
  classroom = seq_len(n_class),
  rater = seq_len(n_rater)
)
class_effect <- rnorm(n_class, sd = 1.3)[grid$classroom]
pupil_effect <- rnorm(n_class * n_pupil, sd = 0.6)[
  (grid$classroom - 1) * n_pupil + grid$pupil
]
rater_effect <- rnorm(n_rater, sd = 0.4)[grid$rater]
school <- data.frame(
  classroom = factor(grid$classroom),
  pupil = factor(paste(grid$classroom, grid$pupil, sep = "_")),
  rater = factor(grid$rater),
  score = 10 + class_effect + pupil_effect + rater_effect +
    rnorm(nrow(grid), sd = 0.7)
)
```

```{r ml-fit, eval = requireNamespace("glmmTMB", quietly = TRUE)}
icc(school, score, subject = pupil, rater = rater, cluster = classroom, seed = 1)
```

Both levels come back in one call. Here the **cluster-level** ICC is the higher
of the two: raters agree more about which *classrooms* score high than about which
*pupils within a classroom* do — exactly the pattern you would expect when most of
the true variation lives between classrooms. Which number you report depends on the
decision you will make: a classroom-level intervention cares about the cluster-level
reliability, a pupil-level one about the subject level. Request just one with
`level = "subject"` or `level = "cluster"`.

## How much does ignoring the nesting cost? The conflated ICC

What would you have reported if you had *ignored* the classrooms and run an ordinary
single-level ICC? That number — ten Hove et al.'s Equation 14 — folds the
between-classroom and within-classroom variation together into one "true score" and
is biased for **both** questions above. `icc()` can compute it as a **diagnostic
contrast** with `level = "conflated"`, so you can see the distortion directly rather
than take it on faith:

```{r ml-conflated, eval = requireNamespace("glmmTMB", quietly = TRUE)}
icc(school, score,
  subject = pupil, rater = rater, cluster = classroom,
  level = c("subject", "cluster", "conflated"), seed = 1
)
```

The conflated value lands between the two correct levels and matches neither: it
over- or under-states the reliability of any real decision. It is printed with a
warning label and is **never** a coefficient to report — it exists only to quantify
the cost of ignoring the structure. (It is absolute-agreement only, and needs a
complete, crossed design.)

## When raters are nested

The classroom example above has every rater rate every pupil in every classroom,
so raters are **crossed** with clusters (ten Hove et al.'s Design 1). Two other
layouts are common, and `icc()` **infers which one you have** from the data — you
never declare it:

- **Raters nested in clusters** (Design 2): each classroom has its *own* panel of
  raters. There is then no between-cluster reliability to report — a cluster-level
  ICC needs the *same* raters spanning clusters — so `icc()` returns the subject
  level only.
- **Raters nested in subjects** (Design 3): each pupil is rated by their *own*
  raters. Now systematic rater differences cannot be separated from residual error
  at all, so this is a multilevel *one-way* design: it reports agreement-only
  `ICC(1)` / `ICC(k)`, the clustered analogue of `model = "oneway"`.

Take the same classrooms but give each one its own raters (Design 2):

```{r ml-nested-clusters, eval = requireNamespace("glmmTMB", quietly = TRUE)}
school_d2 <- school
school_d2$rater <- factor(paste(school_d2$classroom, school_d2$rater, sep = "_"))
icc(school_d2, score, subject = pupil, rater = rater, cluster = classroom, seed = 1)
```

The header now reads *raters nested in clusters* and only the subject level comes
back. If instead each *pupil* has their own raters, the design is a multilevel
one-way (Design 3):

```{r ml-nested-subjects, eval = requireNamespace("glmmTMB", quietly = TRUE)}
school_d3 <- school
school_d3$rater <- factor(paste(school_d3$pupil, school_d3$rater, sep = "_"))
icc(school_d3, score, subject = pupil, rater = rater, cluster = classroom, seed = 1)
```

Here the coefficients are labeled `ICC(1)` / `ICC(k)` and `type` no longer
applies — with each pupil's raters unique, there is no rater main effect to keep
in or drop from the error term. A layout that is neither cleanly crossed nor
cleanly nested (some raters shared across clusters, some not) raises an
informative error rather than guessing at a model.

## Incomplete (ragged) multilevel designs

Just as in the single-level case (the [*Choosing an ICC*](choosing-an-icc.html)
article works a connected incomplete design), the **crossed** design (Design 1) does
not need every pupil rated by every rater. The mixed model estimates the variance
components from whatever cells are present, so a ragged classroom design is handled
directly. Drop a fifth of the ratings from the `school` data at random:

```{r ml-incomplete-data}
set.seed(11)
school_ragged <- school[-sample(nrow(school), round(0.2 * nrow(school))), ]
```

At the **subject** level both agreement and consistency come back, and — exactly as
in the single-level incomplete case — `ICC(*,k)` averages over the *effective* number
of ratings per pupil (`k_eff`, the harmonic mean), which is below the full panel size
of `r n_rater`:

```{r ml-incomplete-subject, eval = requireNamespace("glmmTMB", quietly = TRUE)}
icc(school_ragged, score, subject = pupil, rater = rater, cluster = classroom,
  level = "subject", seed = 1)
```

The header now reads *incomplete*, and the report names the effective `k` so the
divisor is never a black box.

At the **cluster** level, the single-rater `ICC(c,1)` is available on ragged data
(it needs no averaging divisor). Request it with `unit = "single"`:

```{r ml-incomplete-cluster, eval = requireNamespace("glmmTMB", quietly = TRUE)}
icc(school_ragged, score, subject = pupil, rater = rater, cluster = classroom,
  level = "cluster", type = "consistency", unit = "single", seed = 1)
```

Two things are deliberately fenced off on incomplete data, each with a clear error
rather than a silently wrong number. The **averaged** cluster-level `ICC(c,k)` is not
yet supported: the effective number of raters behind a ragged *cluster* mean is a
per-cluster quantity still being validated, distinct from the per-pupil `k_eff`. And
when missing cells make the crossing pattern **ambiguous** — some raters happen to
appear in only one classroom, so the design could be read as crossed *or* nested —
`icc()` does not guess; you resolve it by declaring `design = "crossed"` (validated
against the data), or the abort points you at the nested reading.

## Fixed raters in a multilevel design

The multilevel examples so far treat raters as a **random** sample — the recommended
default, which generalizes beyond the raters you happened to use. When the observed
raters *are* the entire population of interest (a fixed panel of examiners, say), pass
`raters = "fixed"`. As in the single-level case, the rater main effect is then the
finite-population variance of *these* raters (McGraw & Wong's Case 3A) rather than a
random-sample variance. On a balanced crossed design both levels come back:

```{r ml-fixed, eval = requireNamespace("glmmTMB", quietly = TRUE)}
icc(school, score, subject = pupil, rater = rater, cluster = classroom,
  raters = "fixed")
```

On this balanced design the fixed-rater coefficients match the random-rater ones **at
both levels**: **consistency** never uses the rater term, so it is identical either way,
and **absolute agreement** coincides because the finite-population rater variance equals
the random-sample estimate when the design is balanced. At the cluster level the
between-rater disagreement in cluster means is that same finite-population term plus the
cluster-by-rater interaction. The subject level genuinely diverges from random on
**incomplete** data, where the rater variance is estimated from unequal cell counts —
and the crossed (Design 1) fixed-rater multilevel design is supported on ragged data too,
at the subject level, exactly as the single-level incomplete case above.

Multilevel support now covers **random** raters on the crossed design (Design 1),
complete or **incomplete**, at the subject level and — for `ICC(c,1)` — the cluster
level; and the nested designs (Designs 2 and 3, subject level), complete **or
incomplete**. **Fixed** raters are supported at the subject level on the crossed design
(complete **and** incomplete), at the **cluster** level on the crossed design (complete
data), and, on complete data, the nested Design 2. What remains open: the averaged
cluster-level `ICC(c,k)` on incomplete data (the per-cluster effective-rater divisor is
still being validated), incomplete *fixed-rater cluster-level* estimation, and incomplete
*fixed-rater nested* designs. Design 3 reports no fixed-rater or cluster-level coefficient
by construction — with raters nested in subjects there is no separable rater effect to
fix, and no crossed-cluster structure to support a cluster mean.

## How many raters? A multilevel D-study

The [D-study](d-studies-and-replicates.html) works on a multilevel fit too: it
projects the **number of raters** at each level, so you can ask "how many raters
would make the *cluster*-level score reliable?" separately from the subject level.
It returns one curve per level (note the `level` column), and `autoplot()` facets
them:

```{r ml-dstudy, eval = requireNamespace("glmmTMB", quietly = TRUE)}
d_study(
  icc(school, score, subject = pupil, rater = rater, cluster = classroom, seed = 1),
  m = c(1, 2, 4, 8)
)
```

Only the **rater** count is projected. The cluster-level coefficient does not average
over subjects (ten Hove et al. 2022, Eq. 13), so "how many subjects per cluster?" is a
*sample-size* question — how precisely you estimate the variance components — not a
reliability projection.

## Visualizing the levels

The `autoplot()` forest plot (see [*D-studies and within-cell
replicates*](d-studies-and-replicates.html#visualizing-a-fit) for the single-level
version) facets a **multilevel** fit by level, so the subject- and cluster-level
coefficients line up for comparison — the same `school` fit from above, whose cluster
level was the higher of the two:

```{r plot-ml, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("glmmTMB", quietly = TRUE), fig.alt = "Forest plot of the school multilevel fit, faceted into subject-level and cluster-level panels, each showing ICC(A,1) and ICC(A,k) with Monte-Carlo intervals."}
library(ggplot2)
autoplot(icc(school, score, subject = pupil, rater = rater, cluster = classroom,
  seed = 1))
```

## Or let the package choose the level

The multilevel *fifth choice* — subject vs. cluster level — is part of the
`choose_icc()` decision helper too; the [*Choosing an ICC*](choosing-an-icc.html)
guide walks the other four axes. Pass the design and it hands back the coefficient to
report and the exact `icc()` call, without fitting anything:

```{r choose-ml}
choose_icc(model = "twoway", multilevel = TRUE, level = "cluster",
  type = "consistency", unit = "single", raters = "random")
```

Because the helper is generated from the **same estimand machinery** as `icc()`, the
emitted call cannot drift from what `icc()` actually computes. In an **interactive**
session you can omit the deciding answers and `choose_icc()` will ask the outstanding
questions one at a time, then resolve.

## References

ten Hove, D., Jorgensen, T. D., & van der Ark, L. A. (2022). Interrater
reliability for multilevel data: A generalizability theory approach.
*Psychological Methods, 27*(4), 650--666.
